---
title: "Data Cleaning Assignment"
author: "Hylaea Miller"
format: html
warning: FALSE
---

## Week 2 - Arctic Shorebird Demographics Network Data Cleaning

Our goal is to:

-   Clean the Water_cover column to transform it into the correct data type and respect expectations for a percentage

-   Clean the Land_cover column to transform it into the correct data type and respect expectations for a percentage

-   Use the relationship that the three cover columns (Snow, Water, Land) should sum up to 100% to infer missing values where possible and recompute the Total_cover column as needed

## Setup

```{r}
#| message: false
# Load library
library(tidyverse)

# File name
datadir_processed <- file.path("data", "processed")
snowsurvey_file <- "snow_cover.csv"


```

## Data Exploration

We start from the already cleaned Snow_cover output generated during class, which contains the corrected snow cover values.

```{r}
# Import the cleaned snow cover data
cover_df <- read_csv(file.path(datadir_processed, snowsurvey_file))

# Inspect column types and structure
glimpse(cover_df)
```

From `glimpse()` we can see that Water_cover and Land_cover are stored as character columns. Since these represent percentages, they should be numeric. We will clean both columns following the same approach used for Snow_cover


## Water_cover Cleaning

### Exploration

We first identify all non-numeric values to understand what needs to be fixed before converting the column to numeric.

```{r}
# Count non-numeric values in Water_cover
cover_df %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```

We find the same symbols seen in Snow_cover that are not numeric. According to the metadata, "a blank or '-' indicates absence of data", so these values will be replaced with NA. Values such as `n/a`, `unk`, 
and `.` are not defined in the metadata and will be treated as NA for consistency. 

### Fix Non-numeric Values

```{r}
# Replace unknown/missing string values with NA
cover_df <- cover_df %>%
  mutate(
    Water_cover = ifelse(
      Water_cover %in% c(".", "-", "n/a", "unk"), NA, Water_cover
    ))
```

Let’s check again if the symbols are gone.
```{r}
# Verify that no non numeric values remain
cover_df %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```

All good. Let’s do the numeric transformation next.

```{r}
# Convert Water_cover to numeric
cover_df <- cover_df %>%
  mutate(Water_cover = as.numeric(Water_cover))
```

### Validate Percentage Range

Since Water_cover represents a percentage, all values must be between 0 and 100. Values outside this range can be considered not correct and are set to `NA`.

```{r}
# Check for values outside the valid percentage range
cover_df %>%
  filter(Water_cover > 100 | Water_cover < 0)
```
Since we identified one column with values greater than 100, let’s fix it

```{r}
# Set out-of-range values to NA
cover_df <- cover_df %>%
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```



## Land_cover Cleaning

We apply the same cleaning steps to the Land_cover column.

### Exploration

```{r}
# Count non-numeric values in Land_cover
cover_df %>%
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```

The same unknown value symbols are present, so we can apply the same replacements as above.

### Fix Non-numeric Values

```{r}
# Replace unknown/missing string values with NA
cover_df <- cover_df %>%
  mutate(
    Land_cover = ifelse(
      Land_cover %in% c(".", "-", "n/a", "unk"), NA, Land_cover))

```

```{r}
# Convert Land_cover to numeric
cover_df <- cover_df %>%
  mutate(Land_cover = as.numeric(Land_cover))
```

### Validate Percentage Range

```{r}
# Check for values outside the valid percentage range
cover_df %>%
  filter(Land_cover > 100 | Land_cover < 0)
```

Since we identified columns with values greater than 100 and smaller than 0, let’s fix it.

```{r}
# Set out-of-range values to NA
cover_df <- cover_df %>%
  mutate(Land_cover = ifelse(Land_cover > 100 | Land_cover < 0, NA, Land_cover))
```

Let’s check again to make sure the conversion worked.
```{r}
# Check again for values outside the valid percentage range
cover_df %>%
  filter(Land_cover > 100 | Land_cover < 0)
```
It worked!!


## Inferring Missing Values for the Total_cover column

Based on the metadata, the columns Snow_cover + Water_cover + Land_cover should always sum to 100. Where exactly one of the three columns is `NA`, we can infer it as 100 - sum(the other two). We cannot guess the values when two or more columns are missing.

Let's first check how many rows fall into each case:

```{r}
# Count rows by number of NAs across the three cover columns
cover_df %>%
  mutate(na_count = is.na(Snow_cover) + is.na(Water_cover) + is.na(Land_cover)) %>%
  count(na_count)
```

There are 96 rows with exactly one `NA` that can be imputed. Rows with 2 or 3 missing values cannot be recovered.

```{r}
# Impute missing values where exactly one cover column is NA
cover_df <- cover_df %>%
  mutate(
    # Count NAs per row
    na_count = is.na(Snow_cover) + is.na(Water_cover) + is.na(Land_cover),

    # Infer Snow_cover if it is the only missing value
    Snow_cover = ifelse(na_count == 1 & is.na(Snow_cover), 100 - Water_cover - Land_cover, Snow_cover),

    # Infer Water_cover if it is the only missing value
    Water_cover = ifelse(na_count == 1 & is.na(Water_cover), 100 - Snow_cover - Land_cover, Water_cover),

    # Infer Land_cover if it is the only missing value
    Land_cover = ifelse(na_count == 1 & is.na(Land_cover), 100 - Snow_cover - Water_cover, Land_cover)
  ) %>%
  # Drop the helper column before saving
  select(-na_count)
```

Check if it worked

```{r}
# Count again the rows by number of NAs across the three cover columns
cover_df %>%
  mutate(na_count = is.na(Snow_cover) + is.na(Water_cover) + is.na(Land_cover)) %>%
  count(na_count)
```
It worked. No rows with exactly 1 NA

### Handle All-zero Rows

The rows where all three cover columns are 0 cannot represent a valid observation, since snow, water, and land together must cover 100% of the survey area. We consider that these rows are most likely unfilled fields rather than true measurements, so we set them to `NA`.

```{r}
# Set all-zero rows to NA across all three cover columns
cover_df <- cover_df %>%
  mutate(
    all_zero = Snow_cover == 0 & Water_cover == 0 & Land_cover == 0,
    Snow_cover  = ifelse(all_zero, NA, Snow_cover),
    Water_cover = ifelse(all_zero, NA, Water_cover),
    Land_cover  = ifelse(all_zero, NA, Land_cover)
  ) %>%
  select(-all_zero)
```

### Recompute Total_cover

Now, we recompute Total_cover from the cleaned columns. The rows where the three columns do not sum to 100 receive NA, because an explicit missing value is better than a wrong total. I chose to preserve the individual cover columns as-is because most of them can still be used independently for analysis even when the other columns are missing.

```{r}
# Recompute Total_cover; set to NA if columns do not sum to 100
cover_df <- cover_df %>%
  mutate(
    Total_cover = Snow_cover + Water_cover + Land_cover,
    Total_cover = ifelse(Total_cover != 100, NA, Total_cover)
  )
```

### Sanity Check

```{r}
# Confirm no non-NA Total_cover values differ from 100
cover_df %>%
  filter(!is.na(Total_cover), Total_cover != 100) %>%
  nrow()
```

All non-NA totals equal 100!!


## Write Output

```{r}
# check if the folder exists
dir.create(datadir_processed, showWarnings = FALSE)

# Write the cleaned data to a CSV file
write_csv(cover_df, file.path(datadir_processed, "all_cover_fixed_HYLAEA.csv"))
```
